\LARGE \bf GPU Acceleration for Fixed Complexity Sphere Decoder in Large MIMO Uplink Systems

### 2015-08-25 FCSD-CUDA project/The final draft of CUDA paper/conf_draft/GPU Acceleration for Fixed Complexity Sphere Decoder in Large MIMO Uplink Systems.tex
update SVR package

% draft mode.
%\documentclass[a4paper, 10pt, conference]{ieeeconf}
\documentclass[letterpaper, 10pt, conference,twoside]{ieeeconf}
%\documentclass[12pt,a4paper,final]{article}
%\IEEEoverridecommandlockouts
%\overrideIEEEmargins
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,ft.tex

@@

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
%\author{\IEEEauthorblockN{Tianpei Chen} and 
%\IEEEauthorblockN{Harry Leib}}
%\IEEEauthorblockA{ID: 25504946\\Department of Electrical and\\Computer Engineering\\
%McGill University\\
%Montreal, Quebec}}
\author{Tianpei Chen and Harry Leib\\
Department of Electrical and Computer Engineering, McGill University, Montreal, Quebec\\
tianpei.chen@mail.mcgill.ca harry.leib@mcgill.ca}
 %Department of Electrical and Computer Engineering\\
 %McGill University\\
% Montreal, Quebec}

@@

\begin{document}
% make the title area
%\begin{spacing}{2.5}
%\date{}
\maketitle
\thispagestyle{fancy} % IEEE模板在\maketitle后会自动声明\thispagestyle{plain}，  
                      % 导致第一页什么都没有。所以得把plain更改为fancy
%\fancyhead{\thepage} % 奇数页右侧，偶数页左侧显示页码 
\lhead{} % 页眉左，需要东西的话就在{}内添加  
\chead{} % 页眉中  
\rhead{} % 页眉右  
\lfoot{} % 页眉左  
\cfoot{} % 页眉中  
\rfoot{} %页眉右，\thepage 表示当前页码  
\renewcommand{\headrulewidth}{0pt} %改为0pt即可去掉页眉下面的横线  
\renewcommand{\footrulewidth}{0pt} %改为0pt即可去掉页脚上面的横线  
\pagestyle{fancy}  
\fancyhead[RO,LE]{\thepage}  
%\rhead{\thepage}
%\thispagestyle{empty}
%\pagestyle{empty}
% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal technology for next generation wireless communication systems. Detection algorithms for large MIMO schemes tend to require long simulation times on conventional computer systems. In order to improve simulation efficiency of decoders in large MIMO systems, we propose to use General Purpose Graphic Processing Units (GPGPU). This paper considers the implementation of a Fixed Complexity Sphere Decoder (FCSD) for large scale MIMO systems using the framework of GPGPU-Computed Unified Device Architecture (CUDA). Significant speedups were formed for large MIMO systems and constellation sizes over conventional CPU implementation while providing the same Bit Error Rate (BER) performance, hence the proposed algorithm provides great potential for GPGPU to accelerate the simulation of large MIMO systems.

\textbf{ Keywords: Large MIMO, fixed complexity sphere decoder (FCSD), general purpose graphic processing units (GPGPU), computed unified device architecture (CUDA)}
\end{abstract}

@@

\section{Introduction}
A present challenge in the wireless communication area is to bridge the sharp contrast between the increasing demand of mobile network speed for the transmission of rich media content with better QoS (Quality of Service), and the shortage of radio frequency spectrum. Multiple Input Multiple Output (MIMO) technology is of immense research interest since it can help solving this problem owing to several advantages it offers. The core idea of MIMO is to use multiple antennas at the transmitting and receiving sides, that provide parallel data pipes, resulting in spatial multiplexing and diversity gains\cite{oestges2010mimo}.

@@

General Purpose Graphic Processing Units (GPGPU) computing is a common trend in industry and research in areas that need massive computations. Compared to a CPU, a GPU can naturally cope with computational intensive tasks, since most of the hardware resources are allocated for data processing and parallel computation, rather than caching and flow control\cite{cook2013cuda}. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. GPU computing has been applied to different research fields such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435}, and finance\cite{grauer2013accelerating}. The Computed unified Device Architecture (CUDA)\cite{cook2013cuda}\cite{nvidia2008programming} is a programming model for GPGPU that was proposed by NIVIDIA and has become the dominant framework for GPGPU programming. It is based on popular programming languages (C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single Instruction Multiple Threads (SIMT) programming model, enables a GPU to implement parallel data processing, and has become a parallel computing platform for different applications.

 \cite{hongyuan2010fixed} presents an interesting implementation of Fixed Complexity Sphere Decoder (FCSD) based on GPU, however only parallel implementation of multiple decoders is considered rather than exploiting the attribute of parallelism of the algorithm itself. In this paper, a GPU based implementation of single FCSD is proposed, which is referred to as CUDA-FCSD. A heterogeneous programming solution for FCSD preprocessing and integrated blocked parallel path searching and decision making are presented. Using this implementation the speedup as well as Bit Error Rate (BER) performance of different sizes of MIMO systems and signal constellations are analyzed. The proposed CUDA-FCSD implementation demonstrates a great potential for large MIMO systems.

The rest of the paper is organized as follows. Section \ref{programming model} briefly introduces CUDA, the GPU architecture and the programming model. In section \ref{GPUFCSD}, we present details of the implementation of CUDA-FCSD. In section \ref{simulation}, the speedup and bit error rate (BER) performance of CUDA-FCSD are considered. Finally, conclusions are drawn in section \ref{conclusion}.    
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.

@@

\section{CUDA Programming Model and GPU Architecture} \label{programming model} 
\subsection{Single Instruction Multiple Thread Model}
The CUDA programming model is a C programming extension with special defined syntax and structure. The serial code is executed on the host (CPU) side, and massive parallel data processing is performed on the device (GPU) side. Such an approach is defined as a heterogeneous programming model. A thread is the basic programming unit of CUDA. A block of cooperative threads form a thread block, where they can synchronize their behavior and communicate. Fig. \ref{figure1} illustrates a simple CUDA memory hierachy model.
\begin{figure}[htb]
\centering
\includegraphics*[width=0.5\textwidth]{CUDA_programming_hierachy.eps}
\caption{ CUDA 3D memory hierachy $5\times 5\times 5$ grid, with each block composed of $5\times 5\times 5$ threads.
The threads can be organized into one-dimensional, two-dimensional or three-dimensional blocks. Similarly, blocks can also be grouped into a one-dimensional, two-dimensional, or three-dimensional grid.
}
\label{figure1}
\end{figure}

Threads are scheduled, managed, created and executed in a group of 32, called warp. The programs that execute on the device side are defined as \textit{kernel} functions. When a \textit{kernel} function is invoked, the thread blocks are partitioned into warps. The threads in one warp execute the same instruction simultaneously. When one warp is waiting for data loading, the other warps begin to execute. Such an approach is called a Single Instruction Multiple Thread (SIMT) model, whereby the behavior of each thread can be branched to achieve thread level parallelism. The GPU uses a massive number of threads to hide the long latency of a global memory, instead of using a large cache as CPU. A CPU core is designed to minimize the latency for each thread at a time, whereas a GPU is designed to handle a large number of concurrent lightweight threads, in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} 
We present the GPU-GeForce GTX 760 microarchitecture in Fig. \ref{figure2} as an example. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SMs), which perform multiply-add arithmetic operations. Furthermore, each SM has the special functional units (SFU) that can perform more complex operations such as trigonometric functions and reciprocal square root. There are five types of memories: registers, shared memory, global memory, constant memory and texture memory. Registers are allocated dynamically and privately to threads, and provide most rapid access speeds. The shared memory, also called on-chip memory, is located on the SMs and can be shared among the threads in one block. Shared memory has low latency and limited capacity. Global memory, which is off-chip has large latency and large capacity. The other two types of memories, constant memory and texture memory, are located off-chip, but cached. Therefore the access speed of these two type of memory space is much faster than that of global memory. The variables invoked by \textit{kernel} functions can be stored in either of the these five types of memories.

Because of the resource limitation of SM, the total number of threads that can be defined in one block is limited. On current CUDA supported GPUs, one block contains up to 1024 threads. Thus the total number of threads ($T_{total}$) we can use is $T_{total}=N_{SM}\times N_{threads/SM} $, where $N_{SM}$ and $N_{threads/SM}$ denote the number of SMs and number of threads per SM.  
\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{High_view_of_CUDA_GPU_microarchitecture.eps}
\caption{ High View of CUDA CPU Microarchitecture
The GeForce GTX 760 has 6 SMs and the maximum number of threads allowed on each SM is 2048. Thus the total number of threads that can work in parallel is $2048\times 6=12288$.
}
\label{figure2}
\end{figure}   
% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.
\section{GPU Based Acceleration of FCSD}\label{GPUFCSD}
\subsection{MIMO System Model}\label{system}
% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.8 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds
%\hfill December 27, 2012

We consider a complex uncoded spatial multiplexing MIMO system with $N_r$ receive and $N_t$ transmit antennas, $N_{r}\geq N_{t}$, over a flat fading channel. Using a discrete time model, $\mathbf{y}\in\mathbb{C}^{N_{r}\times 1}$ is the received symbol vector written as:
\begin{equation}
\mathbf{y}=\mathbf{H}\mathbf{s}+\mathbf{n},   \label{formula 1}
\end{equation}
where $\mathbf{s}\in \mathbb{C}^{N_{t}\times 1}$ is the transmitted symbol vector, with components that are mutually independent and taken from a signal constellation $\mathbb{O}$ (4-QAM, 16-QAM, 64-QAM) of size $M$. All the possible transmitted symbol vectors $\mathbf{s}\in \mathbb{O}^{N_{t}}$, with $\mathbb{E}[\mathbf{s}\mathbf{s}^{H}]=\mathbf{I}_{N_t}E_{s}$, where $E_{s}$ denotes the symbol average energy, and $\mathbb{E}[\cdot]$ denotes the expectation operation. Furthermore $\mathbf{H}\in \mathbb{C}^{N_{r}\times N_{t}}$ denotes the Rayleigh fading channel propagation matrix with independent identity distributed (i.i.d) circularly symmetric complex Gaussian components of zero mean and unit variance. Finally, $\mathbf{n}\in \mathbb{C}^{N_{r}\times 1}$ is the additive white Gaussian noise (AWGN) vector with zero mean components and $\mathbb{E}[\mathbf{n}\mathbf{n}^{H}]=\mathbf{I}_{N_{r}}N_{0}$, where $N_{0}$ denotes the noise power spectrum density, and hence $\frac{E_{s}}{N_{0}}$ is the signal to noise ratio (SNR).

Assume the receiver has the perfect channel state information (CSI), meaning that $ \mathbf{H}$ is known by receiver, as well as the SNR. The task of the MIMO decoder is to recover $\mathbf{s}$ based on $\mathbf{y}$ and $\mathbf{H}$.
\subsection{Fixed Complexity Sphere Decoder}
From (\ref{formula 1}), the maximum likelihood detector (MLD) for a MIMO system is specified by:
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{y}-\mathbf{H}\mathbf{s}||^{2}. \label{formula 2}
\end{equation}
We consider a MIMO system with  $N_{t}=N_{r}$. Performing the QR decomposition of the channel propagation matrix
\begin{equation}
 \mathbf{H}=\mathbf{Q}\mathbf{R},  \label{QR}
\end{equation}
where $\mathbf{Q}\in \mathbb{C}^{N_{t}\times N_{t}}$ is a unitary matrix and $\mathbf{R}\in \mathbb{C}^{N_{t}\times N_{t}}$, is an upper triangular matrix \cite{golub2012matrix}, we can write (\ref{formula 2}) as:
\begin{eqnarray}
\nonumber
\mathbf{s}_{ML}&=&arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{Q}^{H}\mathbf{H}\mathbf{s}||^{2}\\
&=& arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{R}\mathbf{s}||^{2}. \label{formula 3}
\end{eqnarray}
Using the unconstrained estimator of $\mathbf{s}$, $\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}$, $\mathbf{G}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}$, and (\ref{QR}), we have  
\begin{equation}
\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}=(\mathbf{R}^{H}\mathbf{R})^{-1}\mathbf{R}^{H}\mathbf{Q}^{H}\mathbf{y}
=\mathbf{R}^{-1}\mathbf{Q}^{H}\mathbf{y}.
\label{unconstrained estimation}
\end{equation} 
Therefore $\mathbf{Q}^{H}\mathbf{y}=\mathbf{R}\mathbf{\hat{s}}$, and (\ref{formula 3}) can be written as
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s})||^{2}. \label{formula 4}
\end{equation}

Similarly to SD the FCSD also performs depth first searching. At the first $\rho$ node levels\footnote{when $N_{r}=N_{t}$ the FCSD can achieve the same diversity as the MLD if $\rho\geq \sqrt{N_{t}}-1$ \cite{barbero2008fixing}} the FCSD searches all the possible signal symbols in constellation $\mathbb{O}^{N_{t}}$, a process called the Full Expansion (FE) stage. Let $\mathbf{s}^{F}$ denote FCSD symbol vector candidates. For the symbols $s^{F}_{N_{t}}, s^{F}_{N_{t}-1}\dots s^{F}_{N_{t}-\rho+1}$, the FCSD has $M$ branch expansions at each symbol node. As shown in Fig. \ref{figure4}, there are $\rho=2$ FE node levels and each node level has $M=4$ branches. According to (\ref{formula 4}), we have: 
\begin{equation}
s^{F}_{i}=arg\min_{s\in \mathbb{O}^{N_{t}}}\{\sum_{i=1}^{N_{t}}|r_{ii}(\hat{s}_{i}-s^{F}_{i})+\sum_{j=i+1}^{N_{t}}r_{ij}(\hat{s}_{j}-s^{F}_{j})|^{2}\},  \label{formula 5}
\end{equation} 
$i\in [1,2\dots N_{t}]$, where $r_{ij}$ denotes the component at the $i$th row and $j$th column of the upper triangular matrix $\mathbf{R}$. Based on (\ref{formula 5}) at the remaining $N_{t}-\rho$ levels of symbols, the FCSD employs decision feedback, and therefore there are only one branch expansion for each symbol node at this stage. This stage is called Single Expansion (SE). 
Thus the total number of the branches is fixed, and the FCSD works as a constant number of multiple path tree searching algorithm.

>>>>

% draft mode.
%\documentclass[a4paper, 10pt, conference]{ieeeconf}
\documentclass[letterpaper, 10pt, conference, twoside]{ieeeconf}
%\documentclass[12pt,a4paper,final]{article}
%\IEEEoverridecommandlockouts
%\overrideIEEEmargins
%\documentclass[conference]{IEEEtrans}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,ft.tex

@@

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
%\author{\IEEEauthorblockN{Tianpei Chen} and 
%\IEEEauthorblockN{Harry Leib}}
%\IEEEauthorblockA{ID: 25504946\\Department of Electrical and\\Computer Engineering\\
%McGill University\\
%Montreal, Quebec}}
\author{Tianpei Chen and Harry Leib\\
Department of Electrical and Computer Engineering,\\
 McGill University\\ Montreal, Quebec, Canada\\
tianpei.chen@mail.mcgill.ca harry.leib@mcgill.ca}
 %Department of Electrical and Computer Engineering\\
 %McGill University\\
% Montreal, Quebec}

@@

\begin{document}
% make the title area
%\begin{spacing}{2.5}
%\date{}
\maketitle
%\thispagestyle{fancy} % IEEE模板在\maketitle后会自动声明\thispagestyle{plain}，  
                      % 导致第一页什么都没有。所以得把plain更改为fancy
%\fancyhead{\thepage} % 奇数页右侧，偶数页左侧显示页码 
%\lhead{} % 页眉左，需要东西的话就在{}内添加  
%\chead{} % 页眉中  
%\rhead{} % 页眉右  
%\lfoot{} % 页眉左  
%\cfoot{} % 页眉中  
%\rfoot{} %页眉右，\thepage 表示当前页码  
%\renewcommand{\headrulewidth}{0pt} %改为0pt即可去掉页眉下面的横线  
%\renewcommand{\footrulewidth}{0pt} %改为0pt即可去掉页脚上面的横线  
%\pagestyle{fancy}  
%\fancyhead[RO,LE]{\thepage}  
%\rhead{\thepage}
%\thispagestyle{empty}
%\pagestyle{empty}
% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Large MIMO constitutes a principal technology for next generation wireless communication systems. Detection algorithms for large MIMO schemes tend to require long simulation times on conventional computer systems. In order to improve simulation efficiency of decoders in large MIMO systems, we propose to use General Purpose Graphic Processing Units (GPGPU). This paper considers the implementation of a Fixed Complexity Sphere Decoder (FCSD) for large scale MIMO uplink systems using the framework of GPGPU-Computed Unified Device Architecture (CUDA). Using GPGPU, significant speedups were found for large MIMO systems and constellation sizes, over conventional CPU implementation, while providing the same Bit Error Rate (BER) performance. Hence the proposed approach shows great potential of GPGPU to accelerate computer simulations of large MIMO systems.

\textbf{ \textit{ Keywords: Large MIMO, fixed complexity sphere decoder (FCSD), general purpose graphic processing units (GPGPU), computed unified device architecture (CUDA)}}
\end{abstract}

@@

\section{Introduction}
A present challenge in the wireless communication area is to bridge the sharp gulf between the increasing demand of mobile network speed for the transmission of rich media content with better QoS (Quality of Service), and the shortage of radio frequency spectrum. Multiple Input Multiple Output (MIMO) technology is of immense research interest since it can help solving this problem owing to several advantages it offers. The core idea of MIMO is to use multiple antennas at the transmitting and receiving sides, that provide parallel data pipes, resulting in spatial multiplexing and diversity gains\cite{oestges2010mimo}.

@@

General Purpose Graphic Processing Units (GPGPU) computing is a common trend in industry and research in areas that need massive computations. Compared to a CPU, a GPU can naturally cope with computational intensive tasks, since most of the hardware resources are allocated for data processing and parallel computation, rather than caching and flow control\cite{cook2013cuda}. In recent years GPU computing has provided significant advantages in data processing speeds over CPU computations\cite{nvidia2008programming}. GPU computing has been applied to different research fields such as computer vision\cite{fung2008using}, computer games\cite{blewitt2013applicability}, signal processing\cite{van2011accelerating}\cite{6671435}, and finance\cite{grauer2013accelerating}. The Computed Unified Device Architecture (CUDA)\cite{cook2013cuda}\cite{nvidia2008programming} is a parallel computing platform and programming model for GPGPU that was proposed by NIVIDIA and has become the dominant framework for GPGPU programming. It is based on popular programming languages (e.g. C, C++, Java, Python, Fortran, Direct Compute), such that the application developers in different areas do not need to learn a new programming language or get to know the hardware details. CUDA employs the Single-instruction, Multiple-threads (SIMT) architecture that enables a GPU to implement parallel data processing, and has become a parallel computing platform for different applications.

 An interesting implementation of a Fixed Complexity Sphere Decoder (FCSD) based on GPU is presented in \cite{hongyuan2010fixed}. However, in this scheme only parallel implementation of multiple decoders is considered rather than exploiting the attribute of parallelism of the decoding algorithm itself. In our paper, a GPU based implementation of a single FCSD is proposed, which is referred to as CUDA-FCSD. A heterogeneous programming solution for FCSD preprocessing, integrated blocked parallel path searching and decision making are presented. Using this implementation the speedup as well as Bit Error Rate (BER) performance of different sizes of MIMO systems and signal constellations are analyzed. The proposed CUDA-FCSD implementation demonstrates a great speedup potential for large MIMO systems.

The rest of this paper is organized as follows. Section \ref{programming model} briefly introduces CUDA and GPU architecture. In section \ref{GPUFCSD}, we present details of the implementation of CUDA-FCSD. In section \ref{simulation}, the speedup and bit error rate (BER) performance of CUDA-FCSD are considered. Finally, conclusions are drawn in section \ref{conclusion}.    
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.

@@

\section{CUDA Programming Model and GPU Architecture} \label{programming model} 
\subsection{Single-instruction, Multiple-thread Architecture}
The CUDA programming model is a C programming extension with special defined syntax and principle. The serial code is executed on the host (CPU) side, and massive parallel data processing is performed on the device (GPU) side. Such an approach is defined as a heterogeneous programming model. A thread is the basic programming unit of CUDA. A set of cooperative threads form a thread block, where they can synchronize their behavior and communicate. Fig. \ref{figure1} illustrates a simple CUDA memory hierachy model.
\begin{figure}[htb]
\centering
\includegraphics*[width=0.5\textwidth]{CUDA_programming_hierachy.eps}
\caption{ CUDA 3D memory hierachy $5\times 5\times 5$ grid, with each block composed of $5\times 5\times 5$ threads.
The threads can be organized into one-dimensional, two-dimensional or three-dimensional blocks. Similarly, blocks can also be grouped into a one-dimensional, two-dimensional, or three-dimensional grids.
}
\label{figure1}
\end{figure}

Threads are scheduled, managed, created and executed in a group of 32, called warp. The programs that execute on the device side are defined as \textit{kernel} functions. When a \textit{kernel} function is invoked, the thread blocks are partitioned into warps. The threads in one warp execute the same instruction simultaneously. When one warp is waiting for data loading, the other warps begin to execute. Such an approach is called a Single-instruction, Multiple-thread (SIMT) architecture, whereby the behavior of each thread can be branched to achieve thread level parallelism. The GPU uses a massive number of threads to hide the long latency of a global memory, instead of using a large cache as CPU. A CPU core is designed to minimize the latency for each thread at a time, whereas a GPU is designed to handle a large number of concurrent lightweight threads, in order to maximize throughput.
\subsection{Mircroarchitecture of GPU} 
We present the GPU-GeForce GTX 760 microarchitecture in Fig. \ref{figure2} as an example. The GPU works as a coprocessor system, consisting of Stream Multiprocessors (SMs), which perform multiply-add arithmetic operations. Furthermore, each SM has the special functional units (SFU) that can perform more complex operations such as trigonometric functions and square root. In a GPU, there are five types of memories: registers, shared memory, global memory, constant memory and texture memory. Registers are allocated dynamically and privately to threads, and provide most rapid access. The shared memory, also called on-chip memory, is located on the SMs and can be shared among the threads in one block. Shared memory has low latency and limited capacity. Global memory, which is off-chip, has large latency and large capacity. The other two types of memories, constant memory and texture memory, are also located off-chip, however they are cached. Therefore the access speed of these two types of memory are much faster than that of global memory. The variables invoked by \textit{kernel} functions can be stored in either of the these five types of memory.

Because of the resource limitation of SMs, the total number of threads that can be defined in one block is limited. On current CUDA supported GPUs, one block contains up to 1024 threads. Thus the total number of threads ($T_{total}$) we can use is $T_{total}=N_{SM}\times N_{threads/SM} $, where $N_{SM}$ and $N_{threads/SM}$ denote the number of SMs and number of threads per SM.  
\begin{figure}[htb]
\centering
\includegraphics[width=0.45\textwidth]{High_view_of_CUDA_GPU_microarchitecture.eps}
\caption{ High View of CUDA CPU Microarchitecture
The GeForce GTX 760 has 6 SMs and the maximum number of threads allowed on each SM is 2048. Thus the total number of threads that can work in parallel is $2048\times 6=12288$.
}
\label{figure2}
\end{figure}   
% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.
\section{GPU Based Acceleration of FCSD}\label{GPUFCSD}
\subsection{MIMO System Model}\label{system}
% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.8 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
%I wish you the best of success.
%\hfill mds
%\hfill December 27, 2012

We consider a complex uncoded spatial multiplexing MIMO system with $N_r$ receive and $N_t$ transmit antennas, $N_{r}\geq N_{t}$, over a flat fading channel. Using a discrete time model, $\mathbf{y}\in\mathbb{C}^{N_{r}\times 1}$ is the received symbol vector written as:
\begin{equation}
\mathbf{y}=\mathbf{H}\mathbf{s}+\mathbf{n},   \label{formula 1}
\end{equation}
where $\mathbf{s}\in \mathbb{C}^{N_{t}\times 1}$ is the transmitted symbol vector, with components that are mutually independent and taken from a signal constellation $\mathbb{O}$ (4-QAM, 16-QAM, 64-QAM) of size $M$. The possible transmitted symbol vectors $\mathbf{s}\in \mathbb{O}^{N_{t}}$, satisfy $\mathbb{E}[\mathbf{s}\mathbf{s}^{H}]=\mathbf{I}_{N_t}E_{s}$, where $E_{s}$ denotes the symbol average energy, and $\mathbb{E}[\cdot]$ denotes the expectation operation. Furthermore $\mathbf{H}\in \mathbb{C}^{N_{r}\times N_{t}}$ denotes the Rayleigh fading channel propagation matrix with independent identically distributed (i.i.d) circularly symmetric complex Gaussian components of zero mean and unit variance. Finally, $\mathbf{n}\in \mathbb{C}^{N_{r}\times 1}$ is the additive white Gaussian noise (AWGN) vector with zero mean components and $\mathbb{E}[\mathbf{n}\mathbf{n}^{H}]=\mathbf{I}_{N_{r}}N_{0}$, where $N_{0}$ denotes the noise power spectrum density, and hence $\frac{E_{s}}{N_{0}}$ is the signal to noise ratio (SNR).

Assume the receiver has perfect channel state information (CSI), meaning that $ \mathbf{H}$ is known, as well as the SNR. The task of the MIMO decoder is to recover $\mathbf{s}$ based on $\mathbf{y}$ and $\mathbf{H}$.
\subsection{Fixed Complexity Sphere Decoder}
From (\ref{formula 1}), the maximum likelihood detector (MLD) for a MIMO system is specified by:
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{y}-\mathbf{H}\mathbf{s}||^{2}. \label{formula 2}
\end{equation}
We consider a MIMO system with  $N_{t}=N_{r}$. Performing the QR decomposition of the channel propagation matrix
\begin{equation}
 \mathbf{H}=\mathbf{Q}\mathbf{R},  \label{QR}
\end{equation}
where $\mathbf{Q}\in \mathbb{C}^{N_{t}\times N_{t}}$ is a unitary matrix and $\mathbf{R}\in \mathbb{C}^{N_{t}\times N_{t}}$ is an upper triangular matrix \cite{golub2012matrix}, we can write (\ref{formula 2}) as:
\begin{eqnarray}
\nonumber
\mathbf{s}_{ML}&=&arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{Q}^{H}\mathbf{H}\mathbf{s}||^{2}\\
&=& arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{R}\mathbf{s}||^{2}. \label{formula 3}
\end{eqnarray}
Using the unconstrained estimator of $\mathbf{s}$, $\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}$, $\mathbf{G}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}$, and (\ref{QR}), we have  
\begin{equation}
\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}=(\mathbf{R}^{H}\mathbf{R})^{-1}\mathbf{R}^{H}\mathbf{Q}^{H}\mathbf{y}
=\mathbf{R}^{-1}\mathbf{Q}^{H}\mathbf{y}.
\label{unconstrained estimation}
\end{equation} 
Therefore $\mathbf{Q}^{H}\mathbf{y}=\mathbf{R}\mathbf{\hat{s}}$, and (\ref{formula 3}) can be written as
\begin{equation}
\mathbf{s}_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s})||^{2}. \label{formula 4}
\end{equation}

Similarly to SD the FCSD also performs depth first searching. At the first $\rho$ node levels\footnote{when $N_{r}=N_{t}$ the FCSD can achieve the same diversity as the MLD if $\rho\geq \sqrt{N_{t}}-1$ \cite{barbero2008fixing}} the FCSD searches all the possible signal symbols in constellation $\mathbb{O}$, a process called the Full Expansion (FE) stage. Let $\mathbf{s}^{F}$ denote FCSD symbol vector candidates. For the symbols $s^{F}_{N_{t}}, s^{F}_{N_{t}-1}\dots s^{F}_{N_{t}-\rho+1}$, the FCSD has $M$ branch expansions at each symbol node. As shown in Fig. \ref{figure4}, there are $\rho=2$ FE node levels and each node level has $M=4$ branches. According to (\ref{formula 4}), The decision feedback is: 
\begin{equation}
s^{D}_{i}=arg\min_{s^{D}_{i}\in \mathbb{O}}\{\sum_{i=1}^{N_{t}}|r_{ii}(\hat{s}_{i}-s^{D}_{i})+\sum_{j=i+1}^{N_{t}}r_{ij}(\hat{s}_{j}-s^{D}_{j})|^{2}\},  \label{formula 5}
\end{equation} 
$i\in [1,2\dots N_{t}]$, where $r_{ij}$ denotes the component at the $i$th row and $j$th column of the upper triangular matrix $\mathbf{R}$, $s^{D}_{i}$ denotes the $i$ th symbol solution of decision feedback. Based on (\ref{formula 5}) at the remaining $N_{t}-\rho$ levels of symbols, the FCSD employs decision feedback, therefore there are only one branch expansion for each symbol node at this stage. This stage is called Single Expansion (SE). 
Thus the total number of the branches is fixed, and the FCSD works as a constant number of multiple path tree searching algorithm.

In conclusion the FCSD symbol vector candidate $s^{F}_{i}=[s_{1},s_{2}\dots s_{N_{t}}]$ can be expressed as:
\begin{equation}
\begin{split}
&FE \quad stage:\\
&{s}_{i}^{F}\in    \mathbb{O}^{\rho},  \quad if\quad i=N_{t},N_{t}-1,\dots N_{t}-\rho+1\\
& SE\quad stage:\\
&{s}_{i}^{F}=\mathbb{Q}[\hat{s}_{i}+\sum_{j=i+1}^{N_{t}}\frac{r_{ij}}{r_{ii}}(\hat{s}_{j}-s_{j}^{F})], \quad if\quad \\
& i = N_{t}-\rho,N_{t}-\rho-1,\dots 2,1 
  \label{FCSD solution}
\end{split}
\end{equation}
where $\mathbb{Q}[\cdot]$ denotes the operation of signal constellation quantization. At the post processing stage, the FCSD compares the Euclidean distance of all the symbol vector candidates
\begin{equation}
E_{metric}=||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s}^{F})||^{2}.
\end{equation} 
The symbol vector candidate with minimum $E_{metric}$ is chosen as the solution. The total number of candidates is $M^{\rho}$.  
\begin{figure}[htb]
\centering
\includegraphics[width=0.5\textwidth, height=8cm]{FCSD_tree_searching.eps}
\caption{Tree searching of 4-QAM FCSD for 8$\times$8 MIMO system. The total number of branches is $4^{2}$}
\label{figure4}
\end{figure}
\subsection{CUDA-FCSD Preprocessing}
A FCSD channel ordering strategy is used in order to avoid error propagation in the serial path searching process. This channel ordering strategy is based on post processing signal to noise ratio\cite{wolniansky1998v}
\begin{equation}
\varphi_{m}=\frac{E_{s}}{\sigma^{2}(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}},  \label{ppsnr}
\end{equation}
where $\varphi_{m}$ denotes the post processing signal to noise ratio of the $m$ th data stream, and $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$ denotes the diagonal elements of the inverse of $\mathbf{H}^{H}\mathbf{H}$. The post processing SNR depends only on $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$. From \cite{barbero2008fixing}, at FE stage the data stream with minimum post processing SNR is detected first, while at the SE stage the data stream with maximum post processing SNR is detected first. In conclusion the FCSD ordering works iteratively and obeys the following rule:
at the $j$th step ($j=1,2\dots N_{t}-1$), $s_{p}$ denotes the symbol to be detected, where 
\begin{equation}
p=\left\lbrace \begin{array}{c}
arg\max_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad FE\quad stage\\
arg\min_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad SE\quad stage   \label{the ordering}
\end{array}\right.  
\end{equation}  
and $\mathbf{H}_{j}\in \mathbb{C}^{N_{r}\times (N_{t}-j+1)}$ denotes the renewed channel matrix with the column corresponding to previous detected data streams removed from $\mathbf{H}_{j-1}$.
